Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2023-03-30 letters/response-ref-1-v1.tex

{\bf Response to Associate Editor}
\begin{enumerate}
\item {\it This paper was reviewed by two referees, who tend to agree in their positive assessments of the value and novelty of your work, and provide some very specific comments for further improvements.}

\vspace{5mm}
We thank the associate editor.  We have addressed the two reviewers' specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.   All major edits are highlighted in blue in order to help the referees know what has changed as part of the major revision.
\vspace{5mm}

\item {\it Expanded literature review on existing modeling approaches. Both referees raise this point, and I agree that the contrast with existing works could be more explicit. The second referee provides some specific suggestions on relevant works.}

\vspace{5mm}
We have added a detailed discussion of related work in Section 1.1, which includes the specific suggestions on relevant works provided by the second referee; specifically, we discuss recent modeling of time-to-event data with functional covariates as well as joint modelling from a functional regression perspective.  We contrast these approaches with our design-based approach which becomes necessary when considering high-frequency sensor processes rather than traditional longitudinal studies as the number of measurement occasions increase from tens to thousands/millions of times per individual.  We connect and contrast our use of subsampling with the general use of subsampling in handling massive data and its recent use in massive data with functional predictors.  As stated by the second referee, our proposed joint model includes a recent history functional linear model which we now link with the previous work by Damla Senturk and others.  This motivates Section 5.2 which provides additional empirical study and discussion of the tuning parameter~$\Delta$.
\vspace{5mm}

@@

\item {\it There appear to be some missing definitions of notation, and some transitions/equations need more details to justify them. The latter could potentially be put in the supplement to save space. The former should be added to the main manuscript to improve clarity.}

\vspace{5mm}
We have addressed the missing notation and helped to justify the transitions/equations.  These additions are highlighted in blue.
\vspace{5mm}

\item {\it As the first reviewer, I am confused by $\varepsilon(t) = 0$ on page 4, line 39, in line with equation (3) and measurement error methodology. I suspect comment 7 from the 2nd reviewer alludes to the same thing. I suspect this may be a notation issue rather than the modeling issue, but either way, it needs to be clarified.}

\vspace{5mm}
We have decided that the point we were trying to raise is non-essential to the current manuscript, so we have removed the discussion of \emph{independent evolution} and its implication that $\varepsilon (t) = 0$ for all $t >0$.  The key issue we were trying to clarify is that the traditional joint model formulation in which the observed health process and time-to-event are conditionally independent given the latent process implies that the hazard function conditional on the entire observed health process depends on values after the current time.  In the case study, for example, this implies that the expected risk of a button press given the entire sensor process will depend on future activity index and electrodermal activity.  This holds unless we assume the measurement error is set to zero.  While we still think it is an important point that should be more well appreciated in the joint modeling literature, the current paper has enough content for this discussion to be omitted.  Moreover, the empirical impact seems to be limited and therefore it is better to make this point in a different manuscript and case study.
\vspace{5mm}

@@

\item {\it Practical considerations, such as window size and smoothing parameter selection, need clarification.}

\vspace{5mm}
We added Section 5.2 to discuss choice of window size; specifically, we consider selection based on Akaike information criteria as part of the simulation study rather than simply the impact of incorrect choice of $\Delta$. Table 4 shows that when inference is sensitive to bandwidth selection, the adaptive tuning of $\Delta$ based on AIC does perform well empirically.  We now use this approach in selecting~$\Delta$ in the case study, and provide a plot of AIC across several choices of $\Delta$ in Figure XX in Appendix XX.
\vspace{5mm}

The smoothing parameter selection is also discussed in Section 4.6; given our connection to off-the-shelf penalized logistic regression, the smoothing parameter selection occurs as part of this implementation.  In the R package we are using, `glmnet', the smoothness parameter is chosen via cross-validation.
\vspace{5mm}

### 2023-03-30 write-up/fda-recurrent.tex
fix the write-up

Recall theoretical results were proven under the assumption that there exists~$b^\star$ such that~$\beta(t) = \phi(t) {\bf b}^\star$. To make this assumption plausible, we set~$K_b$ large enough (but less than~$K_x$ to ensure identifiability) to ensure the spline basis expansion is sufficiently expressive. However, in practice, such a choice of~$K_b$ may lead to overfitting the data. Following~\cite{Goldsmith2011}, we choose the polynomial spline model and set~$\phi(t) = b_0 + b_1 t + b_2 t^2 + \sum_{j=3}^{K_b} (t-\kappa_j)^3$ where~$\{ \kappa_j \}_{j=3}^{K_b}$ are the chosen knots and assume~$\{ b_j \}_{j=3}^{K_b} \sim N(0, \sigma^2 I)$ to induce smoothness on the spline model.  Combining the penalized spline formulation with Lemma~\ref{lemma:logistic} establishes a connection between our approximate score equations and solving a generalized mixed effects logistic regression with offset.  Given the connection with generalized mixed effects models, the inferential machinery to obtain variance-covariance estimates . In this context, we acknowledge another limitation.  Specificaly, penalization may lead to confidence intervals that perform poorly in regions where~$\hat \beta(t)$ is oversmoothed.

@@

We next assess the proposed methodology via a simulation study. Here, we assume each individual is observed over five days where each day is defined on the unit interval~$[0,1]$ with $1000$ equally spaced observation times per day. We define~$X(t)$ at the grid of observations as a mean-zero Gaussian process with covariance
\[
  \Sigma (t, t^\prime) =
  \frac{\sigma^2}{\Gamma (v) 2^{\nu-1}}
  \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)^{\nu}
  K_v \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)
\]
where~$K_v$ is the modified Bessel function of the second kind.  We set~$\nu = 1/2$, $\sigma^2 = 1$, and~$\rho = 0.3$ as well as set~$K_b = K_x = 35$.  For simplicity, we assume~$\Sigma$ is known in computation of the eigendecompositions. Given~$\{ X(t) \}_{0 \leq t \leq 1}$ for a given user-day, we generate event times according a chosen hazard function~$h (t; \theta)$.  To mimic our real data, we set
\[
h(t; \theta) = \exp \left( \theta_0 + \int_{0}^{\Delta} X(t-s) \beta(s) ds \right).
\]
We set~$\Delta$ to mimic a 30-minute window for a $12$-hour day.  We set~$\theta_0 = \log(5/1000)$ to set a baseline risk of approximately 5 events per day. We consider two choices of~$\beta(s)$: (1) $\beta_0 + \exp(- \beta_1 s) $, and (2) $\beta_1  * \sin \left( 2 \pi \frac{s}{\Delta} - \pi/2 \right)$.

We generate $1000$ datasets, each consisting of $500$ user-days.  For a given simulated user-day, we randomly sample non-event times using a Poisson process with rate of a halfhour.  We use the proposed methodology to construct the estimate $\hat \beta_{i,0.5} (t)$ for the $i$th simulated dataset; we then subsample the sampled non-event times with thinning probability $1/2$, $1/4$ and $1/8$.  This results in randomly sampled non-event times given by a Poisson process with rates of an hour, two hours, and four hours.  We can construct the corresponding estimates: $\hat \beta_{i,1} (t)$, $\hat \beta_{i,2} (t)$, and $\hat \beta_{i,4} (t)$ respectively. Subsampling allows us to compare the variance due to subsampling as compared to the variance due to sampling fewer non-event times.

Since we are primarily interested in accuracy, we report the mean integrated squared error (MISE) defined as~$\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \beta(s))^2 ds$ for $j=0.5,1,2,4$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  The MISE is defined in this manner to account for settings where~$\Delta$ is unknown.  Next, let $\bar \beta_{j} (t) = \frac{1}{1000} \sum_{i=1}^{1000} \hat \beta_{i,j} (t)$ denote the average estimate for $j=0.5,1,2,4$.  Then the squared bias is given by $\int_{0}^{\infty} ( \bar \beta_{j} (s) - \beta (s))^2 ds$ and the variance is given by $\frac{1}{1000} \sum_{j=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \bar \beta_j(s))^2 ds$.  The subsampling variance is defined as $\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \hat \beta_{i,0.5} (s))^2 ds$.  Table~\ref{tab:mise} show the MISE decomposed into the variance and squared biased as well as the subsampling variance.  To allow fair comparisons across the two choices of $\beta(s)$, all reported numbers are scaled by the integrated square of the true function $\int_{0}^{\infty} \beta(s)^2 ds$. The average runtime (in seconds) is also reported.

Table~\ref{tab:mise} demonstrates that the variance does increase as the sampling rate decreases.  However, the rate of increase in the MISE is relatively quite low.  In the second case study, for example, the MISE increases by 6\% while the run time is 3 times faster.  In the first case study, the MISE remains roughly constant but the run time is 4.7 times faster.  This highlights the efficiency-computation trade-off.  Extrapolating to a complete data analysis, we can see that the run time will continue to increase significantly with minimal improvement in the mean integrated squared error.

@@

When the window length is too large, i.e., $\Delta > \Delta^\star$, then asymptotically the estimation is unbiased as $\beta(t) \equiv 0$ for $t > \Delta$; however, we incur a penalty in finite samples, especially for settings where the function is far from zero near $t = \Delta$.  We find the MISE increases the absolute error $|\Delta - \Delta^\star|$ increases, while the variance decreases as a function of $\Delta$.  While the MISE increased for $\Delta < \Delta^\star$, the pointwise estimation error remains low for $t < \Delta$.  This does not hold for $\Delta > \Delta^\star$, where instead we see parameter attenuation, i.e., a bias towards zero in the estimates at each $0 < t< \Delta$.  To capture this, we define a partial MISE as $\frac{\Delta}{\tilde \Delta} \int_0^{\tilde \Delta} ( \hat \beta_{i,j} (s) - \hat \beta (s) )^2 ds$ where $\tilde \Delta = \min(\Delta, \Delta^\star)$, which is the MISE on the subset $0<t< \min (\Delta, \Delta^\star)$ and scaled for comparative purposes.

@@

A critical consideration for recent history functional linear models~\cite{KIM20111554} is the size of the bandwidth, $\Delta$.  While above we explore the bias-variance trade-off under misspecification of $\Delta$, here we develop a data-adaptive method for bandwidth selection.  Specifically, given the model complexity does not change as a function of $\Delta$, our proposal is to simple compare AIC across a range of bandwidths and choose that which maximizes the criterion.

Table~\ref{tab:aic_selection} presents AIC-based selection across the 500 simulated datasets analyzed in the prior section. We see markedly distinct behavior of the selection criterion in the two settings.  The AIC-based selection works well for the sinusoidal effect across subsampling rates, but performs poorly in the exponential setting. Recall that the sinusoidal effect (choice 2) is significantly different from 0 at $\Delta^\star$ and then 0 for all $s > \Delta^\star$ while the exponential effect (choice 1) decays slowly towards 0 as $s \to^{-} \Delta$. Therefore, the selection problem under the sinusoidal effect is much easier than under the exponential effect.  Moreover, bias, variance, and mean-integrated squared error are not substantially harmed by misspecification of the bandwidth in the exponential setting.  Thus, we conclude that the AIC-based selection method works well in settings where inference requires estimating the bandwidth accurately, as desired.

\begin{table}[!th]
\begin{tabular}{c | c c c c | c c c c}
& \multicolumn{4}{c}{Exponential} & \multicolumn{4}{c}{Sinusoidal} \\ \hline
\diagbox{$\Delta=$}{Rate} & 4.0 & 2.0 & 1.0 & 0.5 & 4.0 & 2.0 & 1.0 & 0.5 \\ \hline
26 & {\bf 0.38} & {\bf 0.35} & {\bf 0.30} & 0.26 & 0.00 & 0.00 & 0.00 & 0.00 \\
29 & 0.22 & 0.20 & 0.17 & 0.12 & 0.09 & 0.14 & 0.15 & 0.18 \\
32 & 0.11 & 0.12 & 0.13 & 0.18 & {\bf 0.85} & {\bf 0.79} & {\bf 0.76} & {\bf 0.70} \\
35 & 0.13 & 0.13 & 0.14 & 0.15 & 0.05 & 0.06 & 0.06 & 0.09 \\
37 & 0.14 & 0.20 & 0.26 & {\bf 0.28} & 0.01 & 0.01 & 0.02 & 0.03 \\ \hline
\end{tabular}
\caption{Selection of bandwidth when true $\Delta = 32$ minutes.
% SINE EXAMPLE
}
\label{tab:aic_selection}
\end{table}

>>>>

Recall theoretical results were proven under the assumption that there exists~$b^\star$ such that~$\beta(t) = \phi(t) {\bf b}^\star$. To make this assumption plausible, we set~$K_b$ large enough (but less than~$K_x$ to ensure identifiability) to ensure the spline basis expansion is sufficiently expressive. However, in practice, such a choice of~$K_b$ may lead to overfitting the data. Following~\cite{Goldsmith2011}, we choose the polynomial spline model and set~$\phi(t) = b_0 + b_1 t + b_2 t^2 + \sum_{j=3}^{K_b} (t-\kappa_j)^3$ where~$\{ \kappa_j \}_{j=3}^{K_b}$ are the chosen knots and assume~$\{ b_j \}_{j=3}^{K_b} \sim N(0, \sigma^2 I)$ to induce smoothness on the spline model.  Combining the penalized spline formulation with Lemma~\ref{lemma:logistic} establishes a connection between our approximate score equations and solving a generalized mixed effects logistic regression with offset.  Given the connection with generalized mixed effects models, the inferential machinery to obtain variance-covariance estimates.  \textcolor{blue}{As we leverage the standard R package `glmnet', the smoothness parameter is chosen via cross-validation.} In this context, we acknowledge another limitation.  Specifically, penalization may lead to confidence intervals that perform poorly in regions where~$\hat \beta(t)$ is over-smoothed.

@@

We next assess the proposed methodology via a simulation study. Here, we assume each individual is observed over five days where each day is defined on the unit interval~$[0,1]$ with $1000$ equally spaced observation times per day. We define~$X(t)$ at the grid of observations as a mean-zero Gaussian process with covariance
\[
  \Sigma (t, t^\prime) =
  \frac{\sigma^2}{\Gamma (v) 2^{\nu-1}}
  \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)^{\nu}
  K_v \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)
\]
where~$K_v$ is the modified Bessel function of the second kind.  We set~$\nu = 1/2$, $\sigma^2 = 1$, and~$\rho = 0.3$ as well as set~$K_b = K_x = 35$.  For simplicity, we assume~$\Sigma$ is known in computation of the eigendecompositions. Given~$\{ X(t) \}_{0 \leq t \leq 1}$ for a given user-day, we generate event times according a chosen hazard function~$h (t; \theta)$.  To mimic our real data, we set
\[
h(t; \theta) = \exp \left( \theta_0 + \int_{0}^{\Delta} X(t-s) \beta(s) ds \right).
\]
We set~$\Delta$ to mimic a 30-minute window for a $12$-hour day.  We set~$\theta_0 = \log(5/1000)$ to set a baseline risk of approximately 5 events per day. We consider two choices of~$\beta(s)$: (1) $\beta_0 + \exp(- \beta_1 s) $ \textcolor{blue}{which decays to 0 as $s$ approaches $\Delta$ from below}, and (2) $\beta_1  * \sin \left( 2 \pi \frac{s}{\Delta} - \pi/2 \right)$ \textcolor{blue}{which is significantly different from 0 as $s$ approaches $\Delta$ from below.}

We generate $1000$ datasets, each consisting of $500$ user-days.  For a given simulated user-day, we randomly sample non-event times using a Poisson process with rate of every five minutes.  We use the proposed methodology to construct the estimate $\hat \beta_{i,12} (t)$ for the $i$th simulated dataset; we then subsample the sampled non-event times with thinning probability $1/3$, $1/6$, $1/12$, and $1/24$.  This results in randomly sampled non-event times given by a Poisson process with rates of every fifteen minutes, half-hour, hour, and two hours.  We can construct the corresponding estimates: $\hat \beta_{i,4} (t)$, $\hat \beta_{i,2} (t)$, $\hat \beta_{i,1} (t)$, and $\hat \beta_{i,0.05} (t)$ respectively. Subsampling allows us to compare the variance due to subsampling as compared to the variance due to sampling fewer non-event times.

Since we are primarily interested in accuracy, we report the mean integrated squared error (MISE) defined as~$\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \beta(s))^2 ds$ for each~$j$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  The MISE is defined in this manner to account for settings where~$\Delta$ is unknown.  Next, let $\bar \beta_{j} (t) = \frac{1}{1000} \sum_{i=1}^{1000} \hat \beta_{i,j} (t)$ denote the average estimate for $j=0.5,1,2,4$.  Then the squared bias is given by $\int_{0}^{\infty} ( \bar \beta_{j} (s) - \beta (s))^2 ds$ and the variance is given by $\frac{1}{1000} \sum_{j=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \bar \beta_j(s))^2 ds$.  The subsampling variance is defined as $\frac{1}{1000} \sum_{i=1}^{1000} \int_{0}^{\infty} ( \hat \beta_{i,j} (s) - \hat \beta_{i,12} (s))^2 ds$.  Table~\ref{tab:mise} show the MISE decomposed into the variance and squared biased as well as the subsampling variance.  To allow fair comparisons across the two choices of $\beta(s)$, all reported numbers are scaled by the integrated square of the true function $\int_{0}^{\infty} \beta(s)^2 ds$. The average runtime (in seconds) is also reported.

Table~\ref{tab:mise} demonstrates that the variance may increase as the sampling rate decreases.  However, the rate of increase in the MISE is relatively quite low.  In the first setting, for example, the RMISE when sampling every hour versus every five minutes is 1.16 while the run time is 14.8 times faster.  In the second setting, the RMISE remains below or equal to $1$ while the run time is 23.1 times faster when sampling every hour rather than every minute.  This highlights the efficiency-computation trade-off.

\begin{newmaterial}
\begin{rmk}[Time-complexity and run-time of maximum likelihood estimation]
The complexity of logistic regression with $n$ observations and covariates of dimension $d$ is $O(nd)$.  Maximum likelihood estimation can be well approximated by subsampling at a very high rate; therefore, under a 4Hz sensor the time-complexity of the maximum likelihood estimate is $(4 \times 60 \times \tau_i) \cdot n \cdot d$ compared to the expected time-complexity under subsampling with rate $c$ per minute of $(c \times \tau_i) \cdot n \cdot d$.  This means using the sensor at its observation frequency of four times per second will take approximately $2^{13}$ times as long as subsampling at a rate of once every half-hour.  In our first example, using the average run times at each sampling rate to project run time at four times per second yields an approximately 2.0 hour run time.  In our second example, we project a run time at four times per second of 6.2 hours.  In both instances, the relative efficiency gain would be negligible suggesting a huge computational increase for minimal relative information gain.
\end{rmk}
\end{newmaterial}

% Extrapolating to a complete data analysis, we can see that the run time will continue to increase significantly with minimal improvement in the mean integrated squared error.

@@

When the window length is too large, i.e., $\Delta > \Delta^\star$, then asymptotically the estimation is unbiased as $\beta(t) \equiv 0$ for $t > \Delta$; however, we incur a penalty in finite samples, especially for settings where the function is far from zero near $t = \Delta$.  We find the MISE increases as the absolute error $|\Delta - \Delta^\star|$ increases.  While the MISE increased for $\Delta < \Delta^\star$, the pointwise estimation error remains low for $t < \Delta$.  This does not hold for $\Delta > \Delta^\star$, where instead we see parameter attenuation, i.e., a bias towards zero in the estimates at each $0 < t< \Delta$.  To capture this, we define a partial MISE as $\frac{\Delta}{\tilde \Delta} \int_0^{\tilde \Delta} ( \hat \beta_{i,j} (s) - \hat \beta (s) )^2 ds$ where $\tilde \Delta = \min(\Delta, \Delta^\star)$, which is the MISE on the subset $0<t< \min (\Delta, \Delta^\star)$ and scaled for comparative purposes.

@@

While above we explore the bias-variance trade-off under bandwidth misspecification, here we explore a data-adaptive method for bandwidth selection, a critical consideration for recent history functional linear models~\citep{KIM20111554}.  Specifically, given the model complexity does not change as a function of $\Delta$, our modest proposal is to compare AIC across a range of bandwidths and choose that which maximizes the criterion. Table~\ref{tab:aic_selection} presents AIC-based selection across the 500 simulated datasets analyzed in the prior section. We see markedly distinct behavior of the selection criterion in the two settings.  The AIC-based selection works well for the sinusoidal effect across subsampling rates, but performs poorly in the exponential setting. Recall that the sinusoidal effect is significantly different from 0 at $\Delta^\star$ and then 0 for all $s > \Delta^\star$ while the exponential effect decays slowly towards 0 as $s \to^{-} \Delta$. Therefore, the selection problem under the sinusoidal effect is much easier than under the exponential effect.  Moreover, bias, variance, and MISE do not vary substantially under bandwidth misspecification in the exponential setting.  Thus, we conclude that the AIC-based selection method works well in settings where inference depends heavily on estimating the bandwidth accurately.

\begin{table}[!th]
\begin{tabular}{c | c c c c | c c c c}
& \multicolumn{4}{c}{Exponential} & \multicolumn{4}{c}{Sinusoidal} \\ \hline
\diagbox{$\Delta=$}{Rate} & 4.0 & 2.0 & 1.0 & 0.5 & 4.0 & 2.0 & 1.0 & 0.5 \\ \hline
26 & {\bf 0.38} & {\bf 0.35} & {\bf 0.30} & 0.26 & 0.00 & 0.00 & 0.00 & 0.00 \\
29 & 0.22 & 0.20 & 0.17 & 0.12 & 0.09 & 0.14 & 0.15 & 0.18 \\
32 & 0.11 & 0.12 & 0.13 & 0.18 & {\bf 0.85} & {\bf 0.79} & {\bf 0.76} & {\bf 0.70} \\
35 & 0.13 & 0.13 & 0.14 & 0.15 & 0.05 & 0.06 & 0.06 & 0.09 \\
37 & 0.14 & 0.20 & 0.26 & {\bf 0.28} & 0.01 & 0.01 & 0.02 & 0.03 \\ \hline
\end{tabular}
\caption{Rate of selection of $\hat \Delta$ across 500 simulations when true $\Delta = 32$ minutes. Bold: most likely bandwidth given a chosen subsampling rate.
% SINE EXAMPLE
}
\label{tab:aic_selection}
\end{table}
